Finite-state morphological transducers for three Kypchak languages

### 2014-03-23 kipchak-lrec-2013/kipchak-paper.tex
another version
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@51110 72bbbca6-d526-0410-a7d9-f06f51895060

%FIXME: keep revision number up to date
The transducers are available / under development in apertium's subversion repository,\footnote{{\scriptsize \url{https://svn.code.sf.net/p/apertium/svn/languages/}}} in the directories \texttt{apertium-kaz}, \texttt{apertium-tat}, and \texttt{apertium-kum}, and compiled versions may be run at \url{http://turkic.apertium.org/}.  The revision of each module in the subversion repository that the numbers in this paper (stem counts, evaluation, etc.) represent is \texttt{r50547} for Kazakh, \texttt{r50260} for Tatar, and \texttt{r50300} for Kumyk.

@@

The Kazakh morphotactics were originally developed based on the Kyrgyz transducer, which was co-authored by the first author, who is fluent in and has a good linguistic knowledge of Kyrgyz and two native speakers of Kyrgyz. The first author also developed the Kazakh morphotactics who is fluent in and has a good linguistic understanding Kazakh.  The morphotactics of Tatar were developed for the most part by the third author, a native speaker of Tatar, who also worked to polish off the morphotactics of the Kazakh transducer.

@@

In order to create the Kumyk transducer, we approached one of the major parts of speech at a time. We started with the nouns, copying the 
continuation lexica\footnote{A continuation lexicon is a set of morphemes, for example, in the Turkic languages there is a continuation lexicon for `case' which includes the possible case suffixes.} (nominal morphotactics) from the Kazakh transducer. The suffixes were then replaced with the Kumyk suffixes according to the grammar \citep{olmesov2000}. Where the grammar was not explicit regarding a suffix form, we looked the corpus for possible forms and at their contexts. The same process was undertaken for verbs.

@@

\begin{figure}
%\centering
\hspace{-35pt}
%\hspace{-1em}
%    \rotatebox{-90}{
%\hspace{-15pt}
\hspace{-10pt}
\includegraphics[width=0.37\textwidth,bb=50 50 554 770,angle=270]{graphs/kumyk-yoldash-coverage}
%}
%~\\
%~\\

@@

\scalebox{0.81}{
\begin{tabular}{lll}
 \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
\hline
 Құдай Өзінің жаратқандарының  &  Аллаһ Үзе яраткан нәрсәләргә карап, & Аллагь Оьзю яратгъан затлагъа\\
 бәріне қарап, өте жақсы екенін көрді. &  аларның бик яхшы икәнен күрде. & къарап, олар бек яхшы экенин гёрген. \\
\hline
 Құдай\tags{<n><nom>} & Аллаһ\tags{<n><nom>} & Аллагь\tags{<n><nom>}  \\
 Өз\tags{<prn><ref><px3sp><gen>} & Үз\tags{<prn><ref><px3sp><nom>} & Оьз\tags{<prn><ref><px3sp><nom>} \\
 жарат\tags{<v><tv><ger\_past><pl><px3sp><gen>} & ярат\tags{<v><tv><gpr\_past>} & ярат\tags{<v><tv><gpr\_past>} \\
 бәрі\tags{<prn><qnt><px3sp><dat>} & нәрсә\tags{<prn><itg><pl><dat>} & зат\tags{<n><pl><dat>} \\
 қара\tags{<v><tv><gna\_perf>} & кара\tags{<v><tv><gna\_perf>} & къара\tags{<v><tv><gna\_perf>} \\ 
,\tags{<cm>} & ,\tags{<cm>} & ,\tags{<cm>} \\
 ---      & алар\tags{<prn><pers><p3><pl><gen>} & олар\tags{<prn><pers><p3><pl><nom>} \\
 өте\tags{<adv>} & бик\tags{<adv>} & бек\tags{<adv>} \\
 жақсы\tags{<adj>} & яхшы\tags{<adj>} & яхшы\tags{<adj>} \\
 е\tags{<cop><ger\_past><px3sp><acc>} & и\tags{<cop><ger\_past><px3sp><acc>} & э\tags{<cop><ger\_past><px3sp><acc>} \\
 көр\tags{<v><tv><ifi><p3><sg>} & күр\tags{<v><tv><past><p3><sg>} & гёр\tags{<v><tv><past><p3><sg>} \\  
 .\tags{<sent>} & .\tags{<sent>} & .\tags{<sent>} \\
\hline
\end{tabular}
}
\label{table:exoutput}
\caption{An example of the output of each of the morphological transducers for the same 
  sentence (``And God saw every thing that he had made, and, behold, it was 
  very good.'', Genesis 1:31). The sentence may be very roughly glossed as ``God own-his created things-to looking,
  their very good being saw''. The output has been 
  abbreviated to only show the appropriate tag sequence in context, the actual output would give multiple interpretations
  that would require further disambiguation. Refer to Appendix~\ref{app:glossary} for tag descriptions.}

\end{table*}

@@

Types of these phenomena include single characters used to represent multiple phonological forms, loanwords, and acronyms and numerals.  A brief overview of each follows, along with examples and details of our solution.

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 \\
	& News & 3.8M & 92.12 $\pm$ 2.72  \\
	& Religion & 851K & 92.49 $\pm$ 1.66 \\\cline{2-4}
	& Average & -- & 90.07 $\pm$ 1.91 \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 159K & 86.35 $\pm$ 2.17 \\
	& News & 5.2M & 89.75 $\pm$ 0.07 \\
	& Religion & 382K & 91.25 $\pm$ 2.55 \\\cline{2-4}
	& Average & -- & 89.12 $\pm$ 1.60 \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 286K &  91.10 $\pm$ 0.86 \\
	& Religion & 227K & 92.47 $\pm$ 1.03 \\\cline{2-4}
	& Average & -- & 91.78 $\pm$ 0.94 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\vspace{-1.5em}
\end{table}

>>>>

The transducers are available / under development in apertium's subversion repository,\footnote{{\scriptsize \url{https://svn.code.sf.net/p/apertium/svn/languages/}}} in the directories \texttt{apertium-kaz}, \texttt{apertium-tat}, and \texttt{apertium-kum}, and compiled versions may be run at \url{http://turkic.apertium.org/}.  The revision of each module in the subversion repository that the numbers in this paper (stem counts, evaluation, etc.) represent is \texttt{r50547} for Kazakh, \texttt{r50260} for Tatar, and \texttt{r50300} for Kumyk.

@@

The Kazakh morphotactics were originally developed based on the Kyrgyz transducer, which was co-authored by the first author---who is fluent in and has a good linguistic knowledge of both Kyrgyz and Kazakh---and two native speakers of Kyrgyz.  The morphotactics of Tatar were developed for the most part by the second author, a native speaker of Tatar, who also worked to polish off the morphotactics of the Kazakh transducer
%The Kazakh morphotactics were originally developed based on the Kyrgyz transducer, which was co-authored by the first author, who is fluent in and has a good linguistic knowledge of Kyrgyz, along with two native speakers of Kyrgyz. The first author also developed the Kazakh morphotactics who is fluent in and has a good linguistic understanding Kazakh.  The morphotactics of Tatar were developed for the most part by the third author, a native speaker of Tatar, who also worked to polish off the morphotactics of the Kazakh transducer.

@@

In order to create the Kumyk transducer, we approached one of the major parts of speech at a time. We started with the nouns, copying the 
continuation lexica\footnote{A continuation lexicon is a set of morphemes, for example, in the Turkic languages there is a continuation lexicon for `case' which includes the possible case suffixes.} (nominal morphotactics) from the Kazakh transducer. The suffixes were then replaced with the Kumyk suffixes according to the grammar \citep{olmesov2000}. Where the grammar was not explicit regarding a suffix form, we looked the corpus for possible forms and at their contexts. The same process was undertaken for verbs.  Phonological rules were developed as needed.

@@

\begin{figure}
%\centering
%\hspace{-35pt}
%\hspace{-1em}
%    \rotatebox{-90}{
%\hspace{-15pt}
\hspace{-10pt}
\includegraphics[width=0.37\textwidth,bb=50 50 554 770,angle=270]{graphs/kumyk-yoldash-coverage}
%}
%~\\
%~\\

@@

\scalebox{0.81}{
\begin{tabular}{lll}
\toprule
 \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
\midrule
 Құдай Өзінің жаратқандарының  &  Аллаһ Үзе яраткан нәрсәләргә карап, & Аллагь Оьзю яратгъан затлагъа\\
 бәріне қарап, өте жақсы екенін көрді. &  аларның бик яхшы икәнен күрде. & къарап, олар бек яхшы экенин гёрген. \\
\midrule
 Құдай\tags{<n><nom>} & Аллаһ\tags{<n><nom>} & Аллагь\tags{<n><nom>}  \\
 Өз\tags{<prn><ref><px3sp><gen>} & Үз\tags{<prn><ref><px3sp><nom>} & Оьз\tags{<prn><ref><px3sp><nom>} \\
 жарат\tags{<v><tv><ger\_past><pl><px3sp><gen>} & ярат\tags{<v><tv><gpr\_past>} & ярат\tags{<v><tv><gpr\_past>} \\
 бәрі\tags{<prn><qnt><px3sp><dat>} & нәрсә\tags{<prn><itg><pl><dat>} & зат\tags{<n><pl><dat>} \\
 қара\tags{<v><tv><gna\_perf>} & кара\tags{<v><tv><gna\_perf>} & къара\tags{<v><tv><gna\_perf>} \\ 
,\tags{<cm>} & ,\tags{<cm>} & ,\tags{<cm>} \\
 ---      & алар\tags{<prn><pers><p3><pl><gen>} & олар\tags{<prn><pers><p3><pl><nom>} \\
 өте\tags{<adv>} & бик\tags{<adv>} & бек\tags{<adv>} \\
 жақсы\tags{<adj>} & яхшы\tags{<adj>} & яхшы\tags{<adj>} \\
 е\tags{<cop><ger\_past><px3sp><acc>} & и\tags{<cop><ger\_past><px3sp><acc>} & э\tags{<cop><ger\_past><px3sp><acc>} \\
 көр\tags{<v><tv><ifi><p3><sg>} & күр\tags{<v><tv><past><p3><sg>} & гёр\tags{<v><tv><past><p3><sg>} \\  
 .\tags{<sent>} & .\tags{<sent>} & .\tags{<sent>} \\
\bottomrule
\end{tabular}
}
\label{table:exoutput}
\caption{An example of the output of each of the morphological transducers for the same 
  sentence (``And God saw every thing that he had made, and, behold, it was 
  very good.'', Genesis 1:31). The sentence may be very roughly glossed as ``God own-his created things-to looking,
  their very good being saw''. The output has been 
  abbreviated to only show the appropriate tag sequence in context, the actual output would give multiple interpretations
  that would require further disambiguation. Refer to Appendix~\ref{app:glossary} for tag descriptions.}
\vspace{-1.5em}
\end{table*}

@@

Types of these phenomena include single characters used to represent multiple phonological forms, loanwords, and acronyms and numerals.  A brief overview of each follows, along with examples and details of our solutions.

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 \\
	& News & 3.8M & 92.12 $\pm$ 2.72  \\
	& Religion & 851K & 92.49 $\pm$ 1.66 \\\cline{2-4}
	& Average & -- & 90.07 $\pm$ 1.91 \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 159K & 86.35 $\pm$ 2.17 \\
	& News & 5.2M & 89.75 $\pm$ 0.07 \\
	& Religion & 382K & 91.25 $\pm$ 2.55 \\\cline{2-4}
	& Average & -- & 89.12 $\pm$ 1.60 \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 286K &  91.10 $\pm$ 0.86 \\
	& Religion & 227K & 92.47 $\pm$ 1.03 \\\cline{2-4}
	& Average & -- & 91.78 $\pm$ 0.94 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\vspace{-2.5em}
\end{table}
